Demes: a specification for interchanging complex demography

### 2022-04-08 paper.tex
Clean up the pop gen model section

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure is formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissible ranges.
This schema, and its accompanying documentation,
fully describe the entities in the model and their
relationships and the required behaviour of implementations.
Since the schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

Below, we detail
\begin{itemize}
    \item The population genetics model in Section~\ref{sec:appendix-pop-gen-model}
    \item The high-level model specification in Section~\ref{sec:appendix-spec}
    \item The desired behaviour of Demes-format parsers in Section~\ref{sec:appendix-parsers}
    \item And the intended scope of Demes in Section~\ref{sec:appendix-scope}
\end{itemize}

\section{Population genetics model details}\label{sec:appendix-pop-gen-model}

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', by which we mean a grouping of individuals that is convenient
for modelling \citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable for our purposes.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either through ancestor-descendant relationships between the populations,
or through processes of continuous or pulse migration.
No other properties of the populations are specified in the model:
we are concerned only with defining the populations, their sizes, and the
movement of individuals between those populations.

Population and event times are written as units in the past, so that time zero
corresponds to the final generation or ``now'', and event times in the past are
values greater than zero with larger values for events that occur in the more
distant past. By having time units increase into the past, we avoid the need to
choose an arbitrary point in history as ``time zero'', so that instead zero
refers to ``now'' or the final generation of a simulation. A natural
specification for time units is in generations, although other time units are
permitted, such as years, accompanied by the generation time.

Population sizes are given as numbers of individuals, and details
such as ploidy levels are considered external to the model.
We therefore focus on the number of individuals as opposed to the number of genomes.
Sizes and mating system details are specified for each population within
epochs.
Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
	I'm not making an attempt yet to revise that later content--I think other authors should
	weigh in.}

\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}

Within a population, we implicitly assume that allele frequency dynamics can
be described by the Wright-Fisher model.  Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously) and the expected frequency (at birth) of the $i^{th}$ allele currently at frequency $p_i$
one generation in the next generation is $p_iw_i/\bar{w}$, where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions. In this framework,
a forward-time simulation of finite populations is equivalent to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970})
and a backwards-time (coalescent) simulation follows the approximations described in \cite[][chapter 3]{wakeley2008-hd}.
Further, this model assumes "soft" selection \citep{christiansen1975hard}, meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
In the \texttt{Discussion}, we discuss alternative models compatible with this framework and outline the expectations for tools implementing
them.
A plethora of forwards and backwards time simulators currently implement this model (REFS).
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

\krtcomment{We need to revisit the paragraph(s) below.  We do not assume individuals are exchangeable, because we do not
	require that all individuals have the same fitness. See, e.g., p 60 of Wakeley for why fitness == exchangeability
	and why that is required for the standard coalescent.}

Within a population, individuals are assumed to be exchangeable, but there are
also parameters for nonrandom mating, such as selfing or cloning rates, which
give the probability that offspring are generated from one generation to the
next by self-fertilisation or cloning of an individual. Cloning and selfing
rates can take any value between zero and one, and if both are specified then
the selfing rate is the probability an offspring is generated through
self-fertilisation given it was produced sexually (i.e. with probability one
minus the cloning rate). We omit any details regarding selection.
Thus, the history of population size changes does not depend on mean fitnesses within
a population nor on differences in mean fitness between populations.
This indepedence of population size and
selection is often called ``soft" selection \citep{christiansen1975hard}.

List of thoughts to turn into a paragraph:

\begin{enumerate}
    \item Selection parameters may be arbitrarily complex.  Different fitness effects in different demes.
          Variation over time.  Etc..
          The specification would become unwieldy and perhaps limiting.
    \item By allowing selfing and cloning, we allow many "standard" models to be considered with "no or little"
          extra effort.
    \item However, by parameterising selfing and cloning as we have, we are assuming that these properties of
          populations at given times can be specified independently from the genetics.
          In other words, mutations that cause selfing to "come and go" are not considered.
          However, consuming software is free to ignore these parameters and model the genetics within
          the framework of a soft selection model.
    \item We don't view these as limitations. Rather, this is what most people are doing most of the time.
          Just now we can finally try to do this part w/fewer errors.
\end{enumerate}

\aprcomment{If selection is placed on top of the demographic model, this implies
that we are living in a soft selection world. Though sizes could just as easily
be interpreted as carrying capacities in a hard selection model.}
\krtcomment{We are firmly in the soft selection world here.  The connection to hard selection
requires some other strong assumptions that will sometimes be met, and sometimes not.
For example, if the mean fitnesses are approximately equal across demes, then a hard selection
model could be similar to a soft selection model.  Relevant literature to dig up: Felsenstein's review,
which will lead to other stuff.}

\jkcomment{More detail here. Also, why only selfing and cloning, why not selection
coefficients and whatnot?}
\krtcomment{Selection is a can of worms: scaling of diploid genotype fitnesses will differ between tools.
The possibility of "GxE" ("different s for same genotype in different demes") can lead to quadratic complexity.}

@@

This model is, by design, very limited. We do not specify any details of
individuals or of the processes that happen \emph{within} populations,
because such details are necessarily complicated, vary greatly by
application, and any attempt to encompass such a broad range of biological
processes is doomed to failure. An ``individual'' (i.e., a discrete
organism) is an abstraction that should have a well defined interpretation
in most cases.
This narrow focus on individuals and their groupings into populations
should ensure that the Demes standard is stable, and not subject to
continual development as more and more elaborations of basic processes
are added. We could never hope to fully encompass all the details
of every possible simulation (including ploidy,
genome sizes, recombination, mutation and gene conversion rates,
selection coefficients, and so on) and so we clearly demarcate the
% FIXME Not quite true here, we're including selfing and cloning rate
role of the specification by including \emph{none} of them.
Such details can be specified by existing mechanisms used by
simulators. If they vary according to the demography (for example,
we have different recombination maps across populations), then
we can unambiguously define these by referring to the population
identifiers and times defined by the Demes model. Rather than attempting
to embed this information \emph{in} the demographic model, we
can define it externally and refer \emph{to} the demography.

@@

>>>>

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure is formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissible ranges.
This schema, and its accompanying documentation,
fully describe the entities in the model and their
relationships and the required behaviour of implementations.
Since the schema is definitive,
\aprcomment{what exactly do we mean by ``definitive''?}
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

Below we detaail the population genetics model in 
Appendix~\ref{sec:appendix-pop-gen-model},
the high-level, human-readable model specification in
Appendix~\ref{sec:appendix-spec}, demographic model parsers and their
intended behaviour in Appendix~\ref{sec:appendix-parsers}, and the
intended scope of Demes in Appendix~\ref{sec:appendix-scope}.
The extended figures in Appendix~\ref{sec:appendix-figures} provides
additional examples and illustrations of Demes usage.

\section{Population genetics model details}
\label{sec:appendix-pop-gen-model}

In Demes, demographic models consist of one or more interacting populations,
or ``demes'', understood to be a collection of individuals that can be
conveniently modelled using a defined set of rules and parameters
\citep{gilmour_demes_1939,gilmour_deme_1955}.
To avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, with the understanding that the
terms are interchangeable.
A population is defined as some collection of individuals that exists for
some period of time, and has a well-defined size (i.e., number of individuals)
during that time period. Individuals can move between populations
either according to their ancestor-descendant relationships
or through processes of continuous or pulse migrations.
Few other properties of the populations are specified in the model:
we are concerned primarily with defining the populations, their sizes, and the
movement of individuals between those populations.
\aprcomment{previous sentence changed to allow for selfing/cloning}

Population and event times are written as units in the past, so that time zero
corresponds to the final generation or ``now'', and event times in the past are
values greater than zero with larger values corresponding to times in the more
distant past. By having time units increase into the past, we avoid the need to
choose an arbitrary point in history as ``time zero''. A natural
specification for time units is in generations, although other time units are
permitted, such as years, accompanied by the generation time.

Population sizes are given as numbers of individuals, and details
such as ploidy levels are considered external to the model.
We therefore focus on the number of individuals as opposed
to the number of genome copies.
Sizes and mating system details are specified for each population within
epochs.
Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

%\krtcomment{The paragraph(s) below subsume(s)/replace(s) some of what appears in later pargraphs.
%	I'm not making an attempt yet to revise that later content--I think other authors should
%	weigh in.}

%\krtcomment{Aaron -- I have perhaps over-simplified the description of marginal allele frequency dynamics?}

Within a population, we implicitly assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously) and the expected frequency (at birth)
of the $i^{th}$ allele currently at frequency $p_i$
one generation in the next generation is $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \cite[][chapter 3]{wakeley2008-hd}.
Further, this model assumes "soft" selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
% In the \texttt{Discussion}, we discuss alternative models compatible
% with this framework and outline the expectations for tools implementing
% them.
\aprcomment{As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as defining ``carrying capacities''.}
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][\aprcomment{and more\dots?}]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

% \krtcomment{
%     We need to revisit the paragraph(s) below.
%     We do not assume individuals are exchangeable, because we do not
% 	require that all individuals have the same fitness.
%     See, e.g., p 60 of Wakeley for why fitness == exchangeability
% 	and why that is required for the standard coalescent.
% }
% \mhcomment{The `exchangeability' assumption also breaks down under 
% 	different reproductive modes, since the sampling of the alleles
%     affects coalescent times. 
% 	For example, in a highly selfing species the coalescent time for two different alleles 
% 	within an individual is much lower than two alleles from different individuals.
%     I've triedto rewrite the below paragraph to update and clarify
%     the reproduction parameters in line with the discussions on Github.
%     Finally, 'rates' imply events occur as Poisson-like 
% 	processes over continuous time, whereas the values used here instead define the 
% 	\emph{fraction} of reproductions of each type.
%     Shall we rename to selfing/clonal `fractions',
% 	or would that mess up the rest of the schema too much?
% }

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilisation or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilisation while the rest
have parents drawn at random from the previous generation.
Depending on the simulation software, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$).
Simulation software that allow variable rates of selfing are expected to clearly
document the expected behaviour.

By allowing the definition of selfing and cloning probabilities, we allow many 
standard models to be defined.
However, by parameterising selfing and cloning as we have, we assume that these
properties of populations can be specified independently from the genetics.
In other words, mutations that cause selfing probabilities to fluctuate within
an epoch are not considered.
\aprcomment{
    Do we still want to say something along these lines?
    ``However, consuming software is free to ignore these parameters
    and model the genetics within
    the framework of a soft selection model.
    We don't view these as limitations. Rather, this is what most people
    are doing most of the time.
    Just now we can finally try to do this part w/fewer errors.''
}
More details of the mathematical properties of selfing and cloning rates
in a coalescent context can be found in \citet{hartfield2016facsexcoal}.

\mhcomment{If requested
I can quickly draw up a schematic diagram to illustrate these partitions.}

@@

This model is, by design, very limited. We do not specify any further details of
individuals or of the processes that happen \emph{within} populations,
because such details are necessarily complicated, vary greatly by
application, and any attempt to encompass such a broad range of biological
processes is doomed to failure. An ``individual'' (i.e., a discrete
organism) is an abstraction that should have a well defined interpretation
in most cases.
This narrow focus on individuals and their groupings into populations
should ensure that the \Demes\ standard is stable, and not subject to
continual development as more and more elaborations of basic processes
are added. We could never hope to fully encompass all the details
of every possible simulation (including ploidy,
genome sizes, recombination, mutation and gene conversion rates,
selection coefficients, and so on) and so we clearly demarcate the
% FIXME Not quite true here, we're including selfing and cloning rate
role of the specification by including \emph{none} of them.
Such details can be specified by existing mechanisms used by
simulators. If they vary according to the demography (for example,
we have different recombination maps across populations), then
we can unambiguously define these by referring to the population
identifiers and times defined by the Demes model. Rather than attempting
to embed this information \emph{in} the demographic model, we
can define it externally and refer \emph{to} the demography.

@@

\section{Extended data and figures}
\label{sec:appendix-figures}
